Oregon: Zip-Level Wildfire Smoke and Health Outcomes (open in wildfire, not script)

### 2017-04-17 r_scripts/analysis/or_result_zip_smoke.Rmd
Create or_result_zip_smoke.Rmd

---
title: "Oregon: Zip-Level Wildfire Smoke and Health Outcomes for Manuscript"
author: "Jingyang Liu"
date: "April 7, 2017"
output: html_document
---

```{r library calls, include=FALSE}
# loading libraries used
library(tidyverse) # import tidyverse
library(survival) # for conditional logistic regression
library(htmlTable) # table
library(lme4) # random-effects model
library(broom) # broom for tidy data from stats models
library(cowplot) # for multi plot image

# spatial packages
library(ggmap) # map package
library(rgdal) # set coord ref system for shape files
library(rgeos) # need it when converting shape to dataframe
library(maptools) # convert shape to dataframe

@@

This document contains results for the association between wildfire smoke PM~2.5~ and cardiovascular (CVD) and respiratory health outcomes. This particular markdown document focuses on the zip code-level analysis. Note there is a county-level analysis that will be paired with CDC mortality data as well. Herein, I compare different PM~2.5~ estimation methods described in further detail below.

Note, epidemiological methods that Sheryl and I think can improve our estimation and reduce some bias around the point estimates will not be the focus of the manuscript.

```{r data import, include = F, echo = FALSE}
# Set working directory and read in files --------------------------------------
# relative path
# set working directory if you want to run seperate chunks
# setwd("./r_scripts")
getwd()

path <- paste0("../data_new/health_data")
# Infile case-crossover dataframes ---------------------------------------------
# Dataframes made in 'chars_3month_binary_smoke_may2016 script
# resp exacerbations
resp_casecross <- read_csv(paste(path, 
                        "respiratory_may_to_sep_time_strat_casecross.csv", 
                        sep = "/"))

# asthma
asthma_casecross <- read_csv(paste(path, 
                          "asthma_may_to_sep_time_strat_casecross.csv", 
                          sep = "/"))

# copd 
copd_casecross <- read_csv(paste(path, 
                        "copd_may_to_sep_time_strat_casecross.csv", 
                        sep = "/"))

# pneum or bronchitis
pneum_casecross <- read_csv(paste(path, 
                        "copd_may_to_sep_time_strat_casecross.csv",
                        sep="/"))
# acute bronchitis
acute_bronch_casecross <- read_csv(paste(path, 
                                "acute_bronch_may_to_sep_time_strat_casecross.csv",
                                sep = "/"))
# cvd
cvd_casecross <- read_csv(paste(path, 
                      "cvd_may_to_sep_time_strat_casecross.csv",
                      sep="/"))
# arrhythmia
arrhythmia_casecross <- read_csv(paste(path, 
                              "arrhythmia_may_to_sep_time_strat_casecross.csv",
                               sep="/"))
# cerebral vascular
cereb_vas_casecross <- read_csv(paste(path, 
                              "cerbrovas_dis_may_to_sep_time_strat_casecross.csv",
                              sep="/"))
# heart failure
hf_casecross <- read_csv(paste(path, 
                      "heart_failure_may_to_sep_time_strat_casecross.csv", 
                      sep="/"))
# ischemic heart disease
ihd_casecross <- read_csv(paste(path, 
                       "isch_heart_dis_may_to_sep_time_strat_casecross.csv",
                       sep="/"))
# myo infarc
mi_casecross <- read_csv(paste(path, 
                      "myocardial_infarc_may_to_sep_time_strat_casecross.csv", sep="/"))

# broken arm
broken_arm_casecross <- read_csv(paste(path, 
                              "broken_arm_may_to_sep_time_strat_casecross.csv",
                              sep="/"))

# read zip pm for counts by zip
smoke_path <- paste0("../data/Oregon/")

zip_pm <- read_csv(paste0(smoke_path, "zip_pm_to_merge_with_chars.csv"))

```{r, eval = FALSE}
## 
# read locations of fires
fire_loc <- read_csv(paste0(smoke_path, "washington_fires_locations201209.csv"))

# import the zip-level met data files Rish sent
zip_met <- read_csv(paste0(smoke_path, "wa_zip_exp_0710_2012_ryan.csv")) %>% 
  # define date variable; original character looked like: 01JUL2012
  mutate(date = as.Date(date, "%d%B%Y"))

# note 12/4/16: read_csv is pretty slow on pc, fread doesn't import date variable 
# that well. Solution may be to use fread but convert to date to use later

@@

In these comparisons, we examine various methods of smoke/PM~2.5~ estimations and associations with health outcomes using a time-stratified case-crossover study design. Health outcomes for a patient with a primary diagnosis of cardiopulmonary health outcomes and their date of admission  (index time) were identified. We then created counterfactual observations for each patient for the same day of the week for the entire wildfire season (MAy 1st to Sep 30th, 2013). We further limited our analyses to claims that were coded as emergency or urgent visits to eliminate bias from patients going in for elective/planned procedures. For patients who have a index time closer to May 1, 2013 or closer to September 30, 2013, their referent observations before or after these dates will be excluded as I will not be able to assign estimates of PM~2.5~ to these referent observations. However, this is not a big issue for a time-stratified design as other referent values throughout the time period average out.

@@

As for **exposure methods**, there are four main estimation methods for PM~2.5~: WRF-Chem Smoke (which substracts the WRF-Chem no fire emission from WRF-chem). For the WRF-Chem variable with the 'smoke' designator, this is WRF-Chem - WRF-Chem no fire. For Global Regression, Geo-Weighted Regression, and Kriging with 'smk' designator, I subtracted off the 'Background' estimates of smoke, which I believe are the monthly averages of PM~2.5~ for a given grid.

@@

### Time-Series of Smoke PM~2.5~ Ranges by Ecoregion

Figures below are for each smoke PM~2.5~ estimation method. The figures are the daily range (min and max) for each zip code in a specific region of the state that local offices of the Washington Department of Ecology has jurisdiction over. The figure first starts out by showing the map of the state, divided in to eco regions used by the state for organization purposes, and then presents the time-series ranges for WRF-Chem, krig, geo-weighted estimation methods.

```{r smk time series range by eco region, echo = F, results='asis', message=F}

## ???
# creation of map and identifying zipcodes in ecoregions----
# import the regional shapefile downloaded from state website
shp_dir <- paste0('../../data/shape_files/WAECY__Ecology_Regions')
shp_dir <- paste0('./data/shape_files/WAECY__Ecology_Regions')

eco_region <- readOGR(dsn = shp_dir, layer = 'WAECY__Ecology_Regions')
# set coord ref system to WGS84 to match projections of ggmap
eco_region <- spTransform(eco_region, "+proj=longlat +datum=WGS84")

# summary(eco_region) # projection is GRS80, may be off

# ggmap and rgdal needed
# Set base map boundary layer
bbox <- c(-125, 45, -116.1, 49.5)

# call black and white map
washington_map <- get_map(location= bbox, source = 'stamen',
                          maptype = 'toner', crop = T)

# call google terrain map
#wash_map_terrain <- get_map(location = "washington state", source = 'google',
#                             maptype = 'terrain', crop = F)

# fortify washington shape data in to dataframe (need maptools package)
eco_df <- tidy(eco_region, region = "ECY_REGION")

# summary(wash_zip_df)

# creation of eco region map (may change colors)
eco_region_map <- ggmap(washington_map) + 
  geom_polygon(data = eco_region, aes(x=long, y=lat, group = group, 
    fill = group), color = "black", alpha = 0.8) +
  scale_fill_manual(name = "Ecology Regions",
                    values = c("#2c7bb6","#abd9e9","#d7191c","#fdae61"), 
                    labels = c("Southwest", "Northwest", "Central", "Eastern"),
                    guide = guide_legend(title.position = "top", order = 1)) +
  geom_point(data = fire_loc, aes(x = Longitude, y = Latitude, shape = "Fire Locations"), 
             color = "black", fill = "red") +
  scale_shape_manual(name = "", values = 24, 
                     guide = guide_legend(title.position = "top", order = 2)) +
  ggtitle("Washington State Ecology \n Regional Office Jurisdiction") +
  xlab("Longitude") +
  ylab("Latitude") +
  theme(#legend.position = "bottom",
        #legend.box = "horizontal"
    )

print(eco_region_map)

# import washington zip shapefile ----
#zip_path <- paste0('../../data/shape_files/wash_zip_2012_shapefile')
zip_path <- paste0('./data/shape_files/wash_zip_2012_shapefile')

wash_zip_map <- readOGR(dsn = zip_path, layer = "wash_zip_2012_shapefile")

# change projection to match eco region
wash_zip_map <- spTransform(wash_zip_map, "+proj=longlat +datum=WGS84")

#summary(eco_region)
#summary(wash_zip_map)

# check overlay visually 
#plot(eco_region, lwd = 5)
#plot(wash_zip_map, add = T)

# find the eco regions that are over a zipcode
zip_eco_region_overlap <- over(wash_zip_map, eco_region)
# output zip IDs from zip shapefile
ZIPCODE <- as.character(wash_zip_map@data$ZCTA5CE10)
# join eco region ids with zip (can be used to indicate eco region)
zip_eco_region <- cbind(ZIPCODE, zip_eco_region_overlap) %>% 
  mutate(eco_region = ifelse(ECY_REGION == "CRO", "Central",
                      ifelse(ECY_REGION == "ERO", "Eastern",
                      ifelse(ECY_REGION == "NWRO", "Northwestern",
                      ifelse(ECY_REGION == "SWRO", "Southwestern", NA))))) %>%  
  select(-OBJECTID, -ECY_REGION)

# range of PM values for each day in each eco region for each method ----
# find the min, median, and max values for each date
#summary(filter(zip_pm, date == "2012-07-9")) # quick sum of first zip obs 
# zip_pm zipcode needs to be character to join with region
zip_pm$ZIPCODE <- as.factor(zip_pm$ZIPCODE)

# glimpse(zip_pm)
# these joins should really be done in a loop, but...
# each column would need to be specified for each pm val, 
# which I'd have to think about more than just each one manually
# wrf smoke
wrf_range <- zip_pm %>% 
  full_join(zip_eco_region, by = "ZIPCODE") %>%  
  group_by(date, eco_region) %>% 
  summarise(min = min(wrf_smk_pm, na.rm = T), 
            med = median(wrf_smk_pm, na.rm = T), 
            max = max(wrf_smk_pm, na.rm = T)) %>% 
  # variable for small multiples
  mutate(est_method = "wrf_smk")

# krig smoke
krig_range <- zip_pm %>% 
  full_join(zip_eco_region, by = "ZIPCODE") %>%  
  group_by(date, eco_region) %>% 
  summarise(min = min(krig_smk_pm, na.rm = T), 
            med = median(krig_smk_pm, na.rm = T), 
            max = max(krig_smk_pm, na.rm = T)) %>% 
  mutate(est_method = "krig_smk")

# geo smoke
geo_range <- zip_pm %>% 
  full_join(zip_eco_region, by = "ZIPCODE") %>%  
  group_by(date, eco_region) %>% 
  summarise(min = min(geo_smk_pm, na.rm = T), 
            med = median(geo_smk_pm, na.rm = T), 
            max = max(geo_smk_pm, na.rm = T)) %>% 
  mutate(est_method = "geo_smk")

# loop plots though

file_name <- c('wrf', 'krig', 'geo')
title_name <- c('WRF-Chem Smoke', 'Kriging Smoke', 'GWR Smoke')
ts_list <- list(wrf_range, krig_range, geo_range)

for(i in 1:3){

df_range <- ts_list[[i]]  
gg_title <- title_name[i]

# create plot for method range
plot_print <- ggplot(df_range, aes(x = date)) +
  geom_ribbon(aes(ymin = min, ymax = max, group = eco_region, 
                  fill=eco_region), alpha = 0.7) +
  scale_fill_manual(name = "Washington State Ecology \n Regional Office Jurisdiction",
                    values = c("#d7191c", "#fdae61", "#abd9e9", "#2c7bb6"),
                    labels = c("Central", "Eastern","Northwest", "Southwest"),
                    guide = F) +
  ylim(0,500) +
  #geom_line(aes(y=med, group=eco_region, color=eco_region)) +
  ggtitle(gg_title) +
  ylab(expression('Smoke PM2.5 µg/m'^3)) +
  xlab('Date') + 
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.y = element_text(angle = 90, size = 12),
    axis.title.x = element_text(size = 14), 
    strip.background = element_rect(fill = 'white'))

# file name
name <- file_name[i]

assign(paste0(name,"_plot"), plot_print)

# Layout needs to be edited and I would like to change colors
# but this should be good enough for a review by CDC
map_timeseries <- ggdraw() +
  draw_plot(eco_region_map, x = 0, y = 0.5, width = 0.5, height = 0.5) +
  draw_plot(wrf_plot, x = 0.5, y = 0.5, width = 0.5, height = 0.5) +
  draw_plot(krig_plot, x = 0, y = 0, width = 0.5, height = 0.5) +
  draw_plot(geo_plot, x = 0.5, y = 0, width = 0.5, height = 0.5)

plot(map_timeseries)
# save 6 inch by 4 inch
ggsave("./fig1.pdf", plot = map_timeseries, 
       width = 12, height = 8, units = "in")

## Days of Smoke In Washington

Here is a figure of the days from May 1st to September 30 2013 where PM~2.5~ exceeded >10.

```{r smoke map, message = F, warning = F, echo = F, results='asis'}

# fire location has a weird date format, but I think is only september

# zip code count of smoke ------------------------------------------------------
zip_smoke_days <- zip_pm %>% 
  # create binary indicators of days with smoke
  mutate(wrf_smoke = if_else(wrf_smk_pm > 0, 1, 0),
         wrf_smoke5 = if_else(wrf_smk_pm > 5, 1, 0),
         wrf_smoke10 = if_else(wrf_smk_pm > 10, 1, 0),
         geo_smoke = if_else(geo_smk_pm > 0, 1, 0),
         geo_smoke5 = if_else(geo_smk_pm > 5, 1, 0),
         geo_smoke10 = if_else(geo_smk_pm > 10, 1, 0)) %>% 
  group_by(ZIPCODE) %>% 
  summarise(wrf_smk_days = sum(wrf_smoke), wrf_smk5_days = sum(wrf_smoke5), 
            wrf_smk10_days = sum(wrf_smoke10), geo_smk_days = sum(geo_smoke), 
            geo_smk_days = sum(geo_smoke), geo_smk5_days = sum(geo_smoke5),
            geo_smk10_days = sum(geo_smoke10)) %>% 
  # change zipcode to character
  mutate(ZIPCODE = as.character(ZIPCODE))

# check smoke days
#summary(zip_smoke_days)

# tidy from broom package now works
wash_zip_df <- tidy(wash_zip_map, region = "ZCTA5CE10") %>% 
  # merge in smoke days
  full_join(zip_smoke_days, by = c("id" = "ZIPCODE"))

# map
print_map <- ggmap(washington_map) + 
  geom_polygon(data = wash_zip_df, 
    aes(x=long, y=lat, group = group, fill = geo_smk10_days), alpha = 0.7) +
  scale_fill_gradient(expression("Number of \nSmoke Days"),
    low = '#2c7bb6', high = '#d7191c') + 
  geom_point(data = fire_loc, aes(x = Longitude, y = Latitude, shape = "Fire Locations"), 
             colour = "black", fill = "red") +
  scale_shape_manual(name = "", values = 24) + 
  xlab("Longitude") +
  ylab("Latitude") +
  # control legend/guide lable order
  guides(fill = guide_colourbar(title.position = "top", order = 1),
         shape = guide_legend(title.position = "top", order = 2))

print(print_map)

ggsave("./fig2.pdf", plot = print_map, 
       width = 12, height = 8, units = "in")

@@

### Background on Time-Stratified Case-Crossover Design

The original time-stratified case-crossover design suggested picking referent periods based on any potential time-varying confounding, *a priori*. For example, if day of the week and the month of the exposure sequence is of concern, one might pick referents in the same day of the week and the same day of the month. The papers that discuss and support the use of a time-stratified design are the Janes et al. 2005 **Epidemiology paper**, *Case-Crossover Analyses of Air Pollution Exposure Data; Referent Selection Strategies and Their Implication for Bias*.

The authors state that with the time-stratified design, the *a priori* selection of the referent period on the same day of the week within the same month has the following benefits:

1. Localizable - Meaning that the likelihood of index times conditional on the referent windows contains information about \beta, which I think means the relative difference in the exposed vs unexposed, conditioned on the exposure (air pollution in their example), is unbiased.

2. Non-ignroable - Meaning that time-varying factors can be ignored when using a conditional logistic regression; controlled through design (matching on day of week within the same season).

More referent periods is better for statistical efficiency, but there is a tradeoff, where additional referent periods may introduce what they call 'overlap bias', which they mathmatically define in Janes 2005 **Statistics in Medicine** paper, *Overlap Bias in the Case-Crossover Design, with Applications to Air Pollution Exposures*. Basically this bias occurs when the exposure assigned in the referent period is conditional on the strategy used to define the referent period.

As Janes et al. state in their 2005 **Statistics in Medicine** paper:

  > *Two of the main strengths of the case-crossover design in the air pollution context are
that it controls fixed confounders and allows for control over time-dependent confounders
by design. Choosing referents that are restricted to the same day of the week and season
as the index time will control for these effects. Though we have not examined the degree
to which various referent selection strategies control bias due to confounding, several points
are clear. Referents that are closer to the index time will achieve greater control of seasonal
confounding. However, sampling referents closer to the index time will also result in a loss
of power due to less heterogeneity in the exposure series, and a smaller number of referents
will also result in lower power.*

I will use both Janes 2005 papers in **Epidemiology** and **Statistics in Medicine** to justify the use of the time-stratified design.

This design has an advantage over other case-crossover designs I've explored in that you can use more referent times, and as the exposure at these times are averaged for each subject over a set time period, this balances out some of the more extreme values of PM~2.5~ that may persist over a couple weeks due to wildfire smoke.

@@

```{r data wrangle with zip met vars, warning = F, echo = F, message = F}

# create dataset that can be merged to case-cross over dataframes
daily_zip_met <- zip_met %>%
  # limit to new mean met variables (getting rid of smoke variables I already have)
  select(1:7) %>% 
  rename(ZIPCODE = ZIP) %>% arrange(ZIPCODE, date) %>%
  group_by(ZIPCODE) %>% 
  # create lag variables
  mutate(
    # absolute temp
    meanAT_lag1 = lag(daily_meanAT, 1, order_by = ZIPCODE), 
    meanAT_lag2 = lag(daily_meanAT, 2, order_by = ZIPCODE),
    meanAT_lag3 = lag(daily_meanAT, 3, order_by = ZIPCODE),
    meanAT_lag4 = lag(daily_meanAT, 4, order_by = ZIPCODE),
    meanAT_lag5 = lag(daily_meanAT, 5, order_by = ZIPCODE),
    # relative humidity
    meanRH_lag1 = lag(daily_meanRH, 1, order_by = ZIPCODE),
    meanRH_lag2 = lag(daily_meanRH, 2, order_by = ZIPCODE),
    meanRH_lag3 = lag(daily_meanRH, 3, order_by = ZIPCODE),
    meanRH_lag4 = lag(daily_meanRH, 4, order_by = ZIPCODE),
    meanRH_lag5 = lag(daily_meanRH, 5, order_by = ZIPCODE),
    # temp
    meanT_lag1 = lag(daily_meanT, 1, order_by = ZIPCODE),
    meanT_lag2 = lag(daily_meanT, 2, order_by = ZIPCODE),
    meanT_lag3 = lag(daily_meanT, 3, order_by = ZIPCODE),
    meanT_lag4 = lag(daily_meanT, 4, order_by = ZIPCODE),
    meanT_lag5 = lag(daily_meanT, 5, order_by = ZIPCODE),
    # precip
    meanPrcp_lag1 = lag(daily_meanPrcp, 1, order_by = ZIPCODE),
    meanPrcp_lag2 = lag(daily_meanPrcp, 2, order_by = ZIPCODE),
    meanPrcp_lag3 = lag(daily_meanPrcp, 3, order_by = ZIPCODE),
    meanPrcp_lag4 = lag(daily_meanPrcp, 4, order_by = ZIPCODE),
    meanPrcp_lag5 = lag(daily_meanPrcp, 5, order_by = ZIPCODE),
    # wind speed
    meanWS_lag1 = lag(daily_meanWS, 1, order_by = ZIPCODE),
    meanWS_lag2 = lag(daily_meanWS, 2, order_by = ZIPCODE),
    meanWS_lag3 = lag(daily_meanWS, 3, order_by = ZIPCODE),
    meanWS_lag4 = lag(daily_meanWS, 4, order_by = ZIPCODE),
    meanWS_lag5 = lag(daily_meanWS, 5, order_by = ZIPCODE)) %>% 
  # sort so all like variables are together
  select(1:3, 8:12, 4, 13:17, 5, 18:22, 6, 23:27, 7, 28:32)

#which(colnames(daily_zip_met)=="meanAT_lag5")
#glimpse(daily_zip_met)

```{r time strat within fire season adj for cdc met, warning = F, echo = F, results='asis'} 
# note above, results = 'asis' needed to print htmlTables

# I removed copd exacerbation and rheuamtoid arthritis

@@

@@

  # extract pm values and divide by 10 and ordered
  which(colnames(df_to_loop)=="background_pm_zip") # code to find column numbers, 91
  pm_estimates_df <- df_to_loop[, c(88, 92:94)]/10  # create 10 unit increases

@@

  print(print_plot)
  # save figure
  ggsave("./fig4.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

# figures for research day
  # plot of only respiratory outcomes

  xtabs(~ combined_point_est_df$outcome)
# subset out global method and respiratory outcomes ----
resp_point_est_df <- combined_point_est_df %>%

  filter(outcome == "All Respiratory" |
         outcome == "Asthma" |
         outcome == "COPD" |
         outcome == "Pneumonia" |
         outcome == "Acute Bronchitis" |
         outcome == "Broken Arm")

## ggplot
  resp_plot <- ggplot(resp_point_est_df, 
                      aes(x = pm_method, y = odds_ratio, colour = pm_method)) +
    # custom color 
    scale_color_manual(name = "Smoke-Estimation Method", 
                       values = c("red", "blue", "#32115C"),
                       guide = guide_legend(title.position = "top", title.hjust = 0.5)) +
    geom_point() + #geom_text(vjust = 0, nudge_x = 0.3) +
    geom_errorbar(aes(ymin=lower95, ymax=upper95), width = 0.2) +
    facet_wrap(~outcome, nrow = 2) +
    geom_hline(yintercept = 1, linetype=2) +
#    ggtitle('Association Between PM2.5 \n from Wildfire Smoke on Hospitalizations') +
    ylab(expression(paste("Odds Ratio for 10µg/m"^3, " Increase in PM"[2.5]))) +
    #xlab('Smoke Estimation Method') +
    # plot theme
    theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"),
    # facet text size
    #strip.text = element_text(size = 8),
    # axis element
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(angle = 90),
    # legend elements
    legend.position = "bottom")
    #legend.text = element_text(size = 8))

  print(resp_plot)

# subset out global method and respiratory outcomes ----
cvd_point_est_df <- combined_point_est_df %>% 
   filter(outcome == "Cardiovascular Disease" |
         outcome == "Arrhythmia" |
         outcome == "Cerebrovascular Disease" |
         outcome == "Heart Failure" |
         outcome == "Ischemic Heart Disease" |
         outcome == "Myocardial Infarction")

## ggplot
  cvd_plot <- ggplot(cvd_point_est_df, 
                      aes(x = pm_method, y = odds_ratio, colour = pm_method)) +
    # custom color 
    scale_color_manual(name = "Smoke-Estimation Method", 
                       values = c("red", "blue", "#32115C"),
                       guide = guide_legend(title.position = "top", title.hjust = 0.5)) +
    geom_point() + #geom_text(vjust = 0, nudge_x = 0.3) +
    geom_errorbar(aes(ymin=lower95, ymax=upper95), width = 0.2) +
    facet_wrap(~outcome, nrow = 2) +
    geom_hline(yintercept = 1, linetype=2) +
#    ggtitle('Association Between PM2.5 \n from Wildfire Smoke on Hospitalizations') +
    ylab(expression(paste("Odds Ratio for 10µg/m"^3, " Increase in PM"[2.5]))) +
    #xlab('Smoke Estimation Method') +
    # plot theme
    theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"),
    # facet text size
    #strip.text = element_text(size = 8),
    # axis element
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(angle = 90),
    # legend elements
    legend.position = "bottom")
    #legend.text = element_text(size = 8))

  print(cvd_plot)

@@

When adjusting for the CDC met variables, the results are slightly attenuated when compared to my adjustments with just WRF-Temp, but my overall conclusions do not change much.

Here is the same table, but within season instead of fire season and adjusting for CDC met variables. Rish is worried about time-variant confounding within a subject that may bias the overall results. This might include seasonal differences in baseline risk of asthma.

### Time-Stratified by Age Categories

Following table and figure look at some outcomes stratified by age category. Note that kids <=15 likely won't have many outcomes like COPD or MI. I have restricted the dataset to just the geo smoke method to reduce the number of rows/points on the tables and figures to make it easier to read for now. Code can easily be changed to have all PM~2.5~ methods.

For some outcomes, there may be some evidence of effect modificaiton, like COPD, where >65 age group is at significant risk. However, the point estimates are relatively similar for the 15 to 65 strata, just not significant. Some error bars are not shown as they are too wide; usually <15 age category where there are only a handful of outcomes.

```{r time stratified age strata cdc met, warning =F, echo = F, results='asis'} 
# note above, results = 'asis' needed to print htmlTables

# For age categories, I want a loop for each age category. 
# Presenting just WRF and GWR

# looking only at wrf chem, kriging, and geo-weighted; taking out 'Global Smoke' method
method_list <- c('WRF-Chem Smoke', 'Kriging Smoke', 'Global Smoke', 'GWR Smoke')

# age category list
age_cat_list <- c(0,1,2)

# create an empty list to row bind dataframes together
datalist1 <- list()
datalist2 <- list()

# dataframe to loop through
  df_to_loop <- data.frame(df_list[i])
  # indication of column
  outcome <- colnames(df_to_loop[3])
  # outcome name
  outcome_name <- outcome_list[i]

  # extract covariates from dataframe
  covariates_df <- df_to_loop[, c(1:16, 27:28, 151:157)]

  # extract pm values and divide by 10 and ordered
  #which(colnames(df_to_loop)=="global_smk_pm_zip") # code to find column numbers
  pm_estimates_df <- df_to_loop[, c(19, 26, 25, 24)]/10  # create 10 unit increases

  # new loop for age categories
  for(k in 0:2){

    # empty matrix (12 x 10 matrix)
    point_estimates <- matrix(nrow = 4, ncol = 10, byrow = T)

    colnames(point_estimates) <- c('outcome', 'age_cat', 'pm_method', 'n', 'n_events', 
                                   'odds_ratio', 'lower95', 'upper95', 'se', 'p_val')

    # fill in the outcome namedataframe before method loop
    point_estimates[, 1] <- outcome_name
    # repeat the age category 4 times for each pm method
    point_estimates[, 2] <- k

    # dataframe for analysis creation
    # bind columns back together 
    df_analysis <- cbind(covariates_df, pm_estimates_df) %>% 
      # remove missing pm values
      filter(!is.na(wrf_smk_pm_zip)) %>% 
      # join in cdc met data
      full_join(daily_zip_met, by = c("ZIPCODE", "date")) %>% 
      # limit to emergency or urgent care
      filter(ADM_TYPE == 1 | ADM_TYPE == 2) %>%
      # limit to specific age category
      filter(age_cat == k) %>% 
      # the following code makes sure that the counterfactual values retained are 
      # symetric in that number of obs before = number of obs after
      mutate(obs_diff_admission = (date - date_admit)/7) 
      # dataframe is already for the entire fire season, so I don't need to subset anymore

  # second loop to run a model for each pm estimation method
    for(j in 26:29){

      # set row number to fill
      row_n <- j-25

      # only run the model if the dataframe has observations
      if(nrow(df_analysis) != 0){
      # conditional logistic regression model
      mod <- clogit(outcome ~ df_analysis[[j]] + daily_meanRH + daily_meanT + daily_meanWS + 
                    daily_meanPrcp + strata(PATIENTID), df_analysis)

      # populate matrix
      point_estimates[row_n, 3] <- method_list[row_n]
      point_estimates[row_n, 4] <- mod$n
      point_estimates[row_n, 5] <- mod$nevent
      # odds ratio
      point_estimates[row_n, 6] <- round(exp(summary(mod)$coefficient[1,1]), 3)

      # 95% lower bound
      point_estimates[row_n, 7] <- round(exp((summary(mod)$coefficient[1,1]) -
                                        1.96*(summary(mod)$coefficient[1,3])), 3)
      # 95% upper bound
      point_estimates[row_n, 8] <- round(exp((summary(mod)$coefficient[1,1]) +
                                        1.96*(summary(mod)$coefficient[1,3])), 3)
      # standard error
      point_estimates[row_n, 9] <- round(summary(mod)$coefficient[1,3], 4)
      # p val
      point_estimates[row_n, 10] <- round(summary(mod)$coefficient[1,5], 4)

>>>>

---
title: "Oregon: Zip-Level Wildfire Smoke and Health Outcomes (open in wildfire, not script)"
author: "Jingyang Liu"
date: "April 7, 2017"
output: html_document
---

@@

This document contains results for the association between wildfire smoke PM~2.5~ and cardiovascular (CVD) and respiratory health outcomes. This particular markdown document focuses on the zip code-level analysis. Herein, I compare different PM~2.5~ estimation methods described in further detail below.

```{r include=FALSE}
library(tidyverse)
library(data.table)
library(survival)
library(htmlTable)
library(readr)
```

```{r data import, include = F, eval = FALSE}

### Limit to only ER visits
## pos = 23
var_list <- c('respiratory', 'asthma', 'pneumonia',  'acute_bronch', 'copd', 
              'cvd', 'isch_heart_dis', 'arrhythmia', 'heart_failure', 
              'cerbrovas_dis', 'myocardial_infarc', 'broken_arm')

N <- NA
n <- 1

for (i in var_list){
  read_path <- paste('../data_new/update/ER/or', i, 'may_to_sep_time_strat_casecross.csv', sep = "_")
  or_disease <- read_csv(read_path)

  or_disease <- or_disease %>% 
    filter(pos=="23") %>%
    # filter the unknown gender
    filter(gender!= "U")

  # Create a permanent case-cross over dataset
  file_name <- paste('../data_new/update/ER/or', i, 'may_to_sep_time_strat_casecross_er.csv', sep = "_")

  # write permanent dataset
  write_csv(or_disease, paste0(file_name))
  N[n] = length(unique(or_disease$personkey))
  n = n+1
}

num_person <- data.frame(rbind(var_list, N))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
## Analysis
path <- paste0("../data_new/update/ER")

resp_casecross <- read_csv(paste(path, 
                                 "or_respiratory_may_to_sep_time_strat_casecross_er.csv", 
                                 sep = "/"))

# asthma
asthma_casecross <- read_csv(paste(path, 
                                   "or_asthma_may_to_sep_time_strat_casecross_er.csv", 
                                   sep = "/"))

# copd 
copd_casecross <- read_csv(paste(path, 
                                 "or_copd_may_to_sep_time_strat_casecross_er.csv", 
                                 sep = "/"))

# pneum or bronchitis
pneum_casecross <- read_csv(paste(path, 
                                  "or_pneumonia_may_to_sep_time_strat_casecross_er.csv",
                                  sep="/"))
 # acute bronchitis
acute_bronch_casecross <- read_csv(paste(path, 
                                         "or_acute_bronch_may_to_sep_time_strat_casecross_er.csv",
                                         sep = "/"))
# cvd
cvd_casecross <- read_csv(paste(path, 
                                "or_cvd_may_to_sep_time_strat_casecross_er.csv",
                                sep="/"))
# arrhythmia
arrhythmia_casecross <- read_csv(paste(path, 
                                       "or_arrhythmia_may_to_sep_time_strat_casecross_er.csv",
                                       sep="/"))
# cerebral vascular
cereb_vas_casecross <- read_csv(paste(path, 
                                      "or_cerbrovas_dis_may_to_sep_time_strat_casecross_er.csv",
                                      sep="/"))
# heart failure
hf_casecross <- read_csv(paste(path, 
                               "or_heart_failure_may_to_sep_time_strat_casecross_er.csv", 
                               sep="/"))
# ischemic heart disease
ihd_casecross <- read_csv(paste(path, 
                                "or_isch_heart_dis_may_to_sep_time_strat_casecross_er.csv",
                                sep="/"))
# myo infarc
mi_casecross <- read_csv(paste(path, 
                               "or_myocardial_infarc_may_to_sep_time_strat_casecross_er.csv", sep="/"))

# broken arm
broken_arm_casecross <- read_csv(paste(path, 
                                       "or_broken_arm_may_to_sep_time_strat_casecross_er.csv",
                                       sep="/"))

# read zip pm for counts by zip
smoke_path <- paste0("../data/Oregon/")
zip_pm <- read_csv(paste0(smoke_path, "zip_pm_to_merge_with_chars.csv"))
```

@@

In these comparisons, we examine various methods of smoke/PM~2.5~ estimations and associations with health outcomes using a time-stratified case-crossover study design. Health outcomes for a patient with a primary diagnosis of cardiopulmonary health outcomes and their date of admission  (index time) were identified. We then created counterfactual observations for each patient for the same day of the week for the entire wildfire season (MAy 1st to Sep 30th, 2013). We further limited our analyses to claims that were coded as emergency or urgent visits to eliminate bias from patients going in for elective/planned procedures. However, we found the filtered population is much larger than the population we chose in Washington, so we choose only the emegency room (hospital) visits. For patients who have a index time closer to May 1, 2013 or closer to September 30, 2013, their referent observations before or after these dates will be excluded as I will not be able to assign estimates of PM~2.5~ to these referent observations. However, this is not a big issue for a time-stratified design as other referent values throughout the time period average out.

@@

As for **exposure methods**, there are three main estimation methods for PM~2.5~: WRF-Chem Smoke (which substracts the WRF-Chem no fire emission from WRF-chem). For the WRF-Chem variable with the 'smoke' designator, this is WRF-Chem - WRF-Chem no fire. For Geo-Weighted Regression, and Kriging with 'smk' designator, I subtracted off the 'Background' estimates of smoke, which I believe are the monthly averages of PM~2.5~ for a given grid. However, the 'Background' is the mean of total period from May 1st to Sep 30th 2013 so I have to assign the mean for every day.

@@

@@

@@

```{r echo = F, warning = F, results='asis'}

df_list <- list(resp_casecross, asthma_casecross, copd_casecross, 
                pneum_casecross, acute_bronch_casecross, cvd_casecross, arrhythmia_casecross,
                cereb_vas_casecross, hf_casecross, ihd_casecross, mi_casecross, 
                broken_arm_casecross)

@@

method_list <- c('WRF-Chem Smoke', 'Kriging Smoke', 'Geo-Weighted Smoke')

@@

  # extract pm values and divide by 10 and ordered
  which(colnames(df_to_loop)=="background_pm_zip") # code to find column numbers, 91
  pm_estimates_df <- df_to_loop[, c(88, 93, 92, 94)]/10  # create 10 unit increases

@@

print(print_plot)
# save figure
ggsave("./fig_only_er.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")
```

@@
